Constructing a universe for the setoid model

### 2021-01-21 paper.tex
Reduced some weird align hackery. Changed aligning to be with respect to the typing colon in several places. Undo commit if not desired.

Finally, we will assume an identity type in the style of Martin-L\"of's
inductive identity type. The main difference is that our identity type is a
$\mProp$-valued relation. We have a transport combinator
\textsf{transp} from which \textsf{J} is derivable.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \textsf{Id} && :\,\, && \{A : \mType \} \to A \to A \to \mProp \\
  & \textsf{refl} && : && \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
  & \textsf{transp} && : && \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1 \\
  & && && \textsf{transp}\ C\ \{x\}\ \{x\}\ e\ u \equiv u
\end{alignat*}

The \textsf{transp} combinator provides a strong elimination principle allowing
to eliminate a strict proposition (the identity type) into arbitrary types.
%
We only use this identity type with such a strong transport in Section~\ref{inductive-universe}.
%
% Note that this identity type must necessarily be assumed as
% primitive; in fact, even though we could have defined \textsf{Id}
% from $W_\mProp$, \textsf{transp} is not definable from $W_\mProp$
% elimination.

@@

In our presentation of the setoid model, contexts are given by setoids, that is,
types together with an equivalence relation. A key point of the model is that the
equivalence relation is valued in $\mProp$ and is thus definitionally proof
irrelevant.
\[
\infer{
  \begin{aligned}
    & | \GG | : \mType \\
    & \GG^\sim : | \GG | \to | \GG | \to \mProp \\
    & \textsf{refl}\ \GG : (\gamma : | \GG |) \to \GG^\sim\,\gamma\,\gamma \\
    & \textsf{sym} \ \GG : \forall\{\gamma_0\ \gamma_1\} \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_0 \\
    & \textsf{trans} \ \GG : \forall\{\gamma_0\ \gamma_1\ \gamma_2\}
    \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_2
    \to \GG^\sim\,\gamma_0\,\gamma_2
  \end{aligned}
}{
  \GG : \Con
}
\]

Types in a context $\GG$ are given by displayed setoids over $\GG$
with a fibration condition given by $\textsf{coe}$, $\textsf{coh}$. In
the following, we sometimes omit implicit quantifications such as the $\forall\{\gamma_0\ \gamma_1\}$ in the type of $\textsf{sym}\,\GG$.
\[
\infer{
  \begin{aligned}
    & | A | : | \GG | \to \mType \\
    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    & \textsf{refl*} : \{\gamma : | \GG | \} (a : | A | \gamma) \to A^\sim\ (\textsf{refl}\ \GG\ \gamma)\ a\ a \\
    & \textsf{sym*} : \forall\{\gamma_0\,\gamma_1\,a_0\,a_1\}\{p : \GG^\sim\ \gamma_0\ \gamma_1\}
    \to A^\sim\ p\ a_0\ a_1 \to A^\sim\ (\textsf{sym}\ \GG\ p)\ a_1\ a_0 \\
    & \textsf{trans*} : A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
    & \textsf{coe} : \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    & \textsf{coh} : (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
    \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

@@

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.
\\
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | \sigma | : | \GG | \to | \GD | \\
    & \sigma^\sim : \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
    & t^\sim : (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]
\end{minipage}
\\

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \blank\sim_\setoidU\blank : \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
\vspace{-0.2em}
\begin{align*}
  & \El : \setoidU \to \mType \\
  & \blank\vdash\blank\sim_\El\blank : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
\end{align*}

Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) \defeq \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \Uty | \defeq \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim \defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty \defeq \reflu \\
  & \ldots \\
  & \textsf{coe}\ \Uty \defeq \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty \defeq \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \EL\ A | \defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim \defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) \defeq \reflel \\
  & \ldots \\
  & \textsf{coe}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\

@@

As in the universe example in Section~\ref{ir-universes}, we now define
$\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \setoidU && : \mType_1 && \El : \setoidU \to \mType \\
  & \setoidU && \defeq \Sigma\ (X : \mType)\ (\inU\ X) \hspace{5em} && \El \defeq \pi_1
\end{alignat*}

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\blank\sim_{\setoidU}\blank$ should be a $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly is not, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  & \blank\sim_{\setoidU}\blank : \setoidU \to \setoidU \to \mProp_1 \\
  & (A , a) \sim_{\setoidU} (B , b) =
   \| \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \|
\end{align*}

We are now left to define the indexed equivalence relation on $\El$:
\vspace{-0.2em}
\begin{align*}
  & \blank\vdash\blank\sim_{\El}\blank : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
  & p \vdash x \sim_{\El} y \defeq \ ?
\end{align*}

@@

Contexts in \textsf{Con} are formed out of empty contexts $\bullet$ and context
extension $\blank,\blank$. Types in \textsf{Ty} are either the base type $\iota$ or
$\Pi$ types.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \bullet && : \textsf{Con}                                                          &&  \iota && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
  & \blank,\blank && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
\end{alignat*}

@@

The definition of the codes and predicate types follows that of the original
inductive-inductive type, and can be derived systematically from it. More
importantly, they can be defined without induction-induction, since although
$\textsf{Con}_0$ and $\textsf{Ty}_0$ are defined mutually, their sorts are not
indexed. \\
\vspace{-0.2em}
\begin{minipage}{0.4\textwidth}
\begin{alignat*}{10}
  & \bullet_0 && : \textsf{Con}_0 \\
  & \blank,_0\blank && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
  & \iota_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \\
  & \Pi_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
\end{alignat*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-1.3em}
\begin{alignat*}{10}
  & \bullet_1 && : \textsf{Con}_1\ \bullet_0 \\
  & \blank,_1\blank && : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
  & && \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
  & \iota_1 && : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
  & \Pi_1 && : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
  & && \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
  & && \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
\end{alignat*}
\end{minipage}
\\

>>>>

Finally, we will assume an identity type in the style of Martin-L\"of's
inductive identity type. The main difference is that our identity type is a
$\mProp$-valued relation. We have a transport combinator
\textsf{transp} from which \textsf{J} is derivable.
\vspace{-0.2em}
\begin{align*}
  \textsf{Id} &: \{A : \mType \} \to A \to A \to \mProp \\
  \textsf{refl} &: \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
  \textsf{transp} &: \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1
\end{align*}
with $\textsf{transp}\ C\ \{x\}\ \{x\}\ e\ u \equiv u$.
The \textsf{transp} combinator provides a strong elimination principle allowing
to eliminate a strict proposition (the identity type) into arbitrary types.
%
We only use this identity type with such a strong transport in Section~\ref{inductive-universe}.
%
% Note that this identity type must necessarily be assumed as
% primitive; in fact, even though we could have defined \textsf{Id}
% from $W_\mProp$, \textsf{transp} is not definable from $W_\mProp$
% elimination.

@@

In our presentation of the setoid model, contexts are given by setoids, that is,
types together with an equivalence relation. A key point of the model is that the
equivalence relation is valued in $\mProp$ and is thus definitionally proof
irrelevant.
\[
\infer{
  \begin{aligned}
    | \GG | &: \mType \\
    \GG^\sim &: | \GG | \to | \GG | \to \mProp \\
    \textsf{refl}\ \GG &: (\gamma : | \GG |) \to \GG^\sim\,\gamma\,\gamma \\
    \textsf{sym} \ \GG &: \forall\{\gamma_0\ \gamma_1\} \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_0 \\
    \textsf{trans} \ \GG &: \forall\{\gamma_0\ \gamma_1\ \gamma_2\}
    \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_2
    \to \GG^\sim\,\gamma_0\,\gamma_2
  \end{aligned}
}{
  \GG : \Con
}
\]

Types in a context $\GG$ are given by displayed setoids over $\GG$
with a fibration condition given by $\textsf{coe}$, $\textsf{coh}$. In
the following, we sometimes omit implicit quantifications such as the $\forall\{\gamma_0\ \gamma_1\}$ in the type of $\textsf{sym}\,\GG$.
\[
\infer{
  \begin{aligned}
    | A | &: | \GG | \to \mType \\
    A^\sim &: \{\gamma_0\ \gamma_1 : | \GG |\}
    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    \textsf{refl*} &: \{\gamma : | \GG | \} (a : | A | \gamma) \to A^\sim\ (\textsf{refl}\ \GG\ \gamma)\ a\ a \\
    \textsf{sym*} &: \forall\{\gamma_0\,\gamma_1\,a_0\,a_1\}\{p : \GG^\sim\ \gamma_0\ \gamma_1\}
    \to A^\sim\ p\ a_0\ a_1 \to A^\sim\ (\textsf{sym}\ \GG\ p)\ a_1\ a_0 \\
    \textsf{trans*} &: A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
    \textsf{coe} &: \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    \textsf{coh} &: (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
    \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

@@

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.
\\
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    | \sigma | &: | \GG | \to | \GD | \\
    \sigma^\sim &: \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    | t | &: (\gamma : | \GG |) \to | A | \, \gamma \\
    t^\sim &: (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]
\end{minipage}
\\

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  \textsf{data} \ \setoidU &: \mType_1 \\
  \blank\sim_\setoidU\blank &: \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  \textsf{refl}_\setoidU &: (A : \setoidU) \to \equ{A}{A} \\
  \textsf{sym}_\setoidU &: \equ{A}{B} \to \equ{B}{A} \\
  \textsf{trans}_\setoidU &: \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
\vspace{-0.2em}
\begin{align*}
  \El &: \setoidU \to \mType \\
  \blank\vdash\blank\sim_\El\blank &: \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
\end{align*}

Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  \textsf{refl}_\El &: (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  \textsf{sym}_\El &: \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  \textsf{trans}_\El &: \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  \coeel &: \equ{A}{B} \to \El\ A \to \El\ B \\
  \cohel &: (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) \defeq \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  | \Uty | &\defeq \lambda\, \gamma . \, \setoidU \\
  \Uty^\sim &\defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  \textsf{refl}\ \Uty &\defeq \reflu \\
  &\ldots \\
  \textsf{coe}\ \Uty &\defeq \lambda\,p\,a . \, a \\
  \textsf{coh}\ \Uty &\defeq \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  | \EL\ A | &\defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  (\EL\ A)^\sim &\defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  \textsf{refl}\ (\EL\ A) &\defeq \reflel \\
  &\ldots \\
  \textsf{coe}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  \textsf{coh}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\

@@

As in the universe example in Section~\ref{ir-universes}, we now define
$\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
\vspace{-0.2em}
\begin{align*}
  \setoidU & : \mType_1 & \El &: \setoidU \to \mType \\
  \setoidU & \defeq \Sigma\ (X : \mType)\ (\inU\ X) & \El &\defeq \pi_1
\end{align*}

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\blank\sim_{\setoidU}\blank$ should be a $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly is not, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  \blank\sim_{\setoidU}\blank &: \setoidU \to \setoidU \to \mProp_1 \\
  (A , a) \sim_{\setoidU} (B , b) &\defeq
   \| \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \|
\end{align*}

We are now left to define the indexed equivalence relation on $\El$:
\vspace{-0.2em}
\begin{align*}
  \blank\vdash\blank\sim_{\El}\blank & : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
  p \vdash x \sim_{\El} y & \defeq \ ?
\end{align*}

@@

Contexts in \textsf{Con} are formed out of empty contexts $\bullet$ and context
extension $\blank,\blank$. Types in \textsf{Ty} are either the base type $\iota$ or
$\Pi$ types.
\vspace{-0.2em}
\begin{align*}
  \bullet &: \textsf{Con}                                                                  &  \iota &: (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
  \blank,\blank &: (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &  \Pi &: \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
\end{align*}

@@

The definition of the codes and predicate types follows that of the original
inductive-inductive type, and can be derived systematically from it. More
importantly, they can be defined without induction-induction, since although
$\textsf{Con}_0$ and $\textsf{Ty}_0$ are defined mutually, their sorts are not
indexed. \\
\vspace{-0.2em}
\begin{minipage}{0.4\textwidth}
\begin{align*}
  \bullet_0 &: \textsf{Con}_0 \\
  \blank,_0\blank &: \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
  \iota_0 &: \textsf{Con}_0 \to \textsf{Ty}_0 \\
  \Pi_0 &: \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-1.3em}
\begin{alignat*}{10}
  \bullet_1 &: \textsf{Con}_1\ \bullet_0 \\
  \blank,_1\blank &: \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
  & \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
  \iota_1 &: \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
  \Pi_1 &: \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
  & \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
  & \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
\end{alignat*}
\end{minipage}
\\
